A note on observation processes in epidemic models

### 2019-07-23 ms/ms.tex
.

@@

Mechanistic analyses of epidemic time series allow us to infer the underlying 
transmission mechanism, estimate biologically relevant parameters, and 
predict the course of an outbreak \citep{breto2009time}. In order to make a 
precise and accurate inference, disease modelers often focus on developing
more realistic process models. For example, the same London measles time 
series from the prevaccination era has been analyzed using multiple models -- 
these models account for time-varying transmission rates \citep{fine1982measles}, 
realistic age structure \citep{schenzle1984age},
metapopulation structure \citep{xia2004measles}, continuous-time infection process 
\citep{cauchemez2008likelihood}, and extra-demographic variability \citep{he2009plug}.

@@

The Susceptible-Infected-Removed (SIR) model describes how disease spreads in a
homogeneous population:
\begin{equation}
\begin{aligned}
\frac{dS}{dt} &= - \beta S \frac{I}{N}\\
\frac{dI}{dt} &= \beta S \frac{I}{N} - \gamma I\\
\frac{dR}{dt} &= \gamma I,
\end{aligned}
\end{equation}
where $\beta$ is the contact rate per unit time, $\gamma$ is the recovery rate per unit time, 
and $N = S + I + R$ is the total population size. 
We define incidence at time $t$ as the number of newly infected
individuals that are infected between time $t- \Delta t$ and time $t$, where $\Delta t$ is
the reporting time step. We expect infected individuals to report their infection some time
after their infection; the number of reported cases defines the \emph{observed} incidence.

For brevity, we consider two extreme cases: individuals instantaneously report
their infection when they become infected or when they recover. The observed incidence 
measured upon infection ($i_1(t)$) can be defined by the integral:
\begin{equation}
i_1(t) = \int_{t - \Delta t}^{t} \beta S \frac{I}{N} dt,
\end{equation}
where $\Delta t$ is the reporting time step.
Alternatively, we can keep track of cumulative incidence, $C$, by adding a 
new state variable described by $dC/dt = \beta S I/N$ and taking the difference between 
the two consecutive reporting periods: $i_1(t) = C(t) - C(t-\Delta t)$. Likewise, 
incidence measured upon recovery ($i_2(t)$) can be defined by the integral:
\begin{equation}
i_2(t) = \int_{t-\Delta t}^{t} \gamma I dt,
\end{equation}
or by the consecutive difference in the cumulative number of recovered cases:
$i_2(t) = R(t) - R(t - \Delta t)$.
Finally, we model under-reporting using a negative binomial distribution with
means $\rho i_1(t)$ and $\rho i_2(t)$, where $\rho$ is the reporting rate, and
an over-dispersion parameter $\theta$. For convenience, we will refer to these two
negative binomial models as infection model and recovery model hereafter; 
likewise, we will refer to epidemic time series generated from these models 
as infection time series and recovery time series.

@@

In order to introduce the problem, we first compare the dynamics of observed 
incidence curves, $i_1(t)$ and $i_2(t)$, (Figure 1). A lag in reporting time delays
the observed epidemic peak timing and reduce the size of the peak. However, 
these differences in the reporting time have small effect to the overall shape 
of the epidemic curve. In the presence of observation and process error, we 
would not expect to be able to distinguish between the two reporting processes.

Note that incidence, regardless of when it is measured, is different from prevalence,
$I(t)$, which is defined as the number of \emph{currently} infected individuals;
the dynamics of observed incidence depend on the reporting time step (because the sum of 
true incidence is equal to the final size of an epidemic), whereas those of
prevalence do not. We expect the dynamics of incidence and prevalence to be similar only when
the reporting time step is equal to the disease generation time. While it is
uncommon, some models do not make a clear distinction between the two (e.g., 
\cite{hooker2010parameterizing}).

@@

Disease modelers often assume that the mean infectious period of a disease
is known and focus on estimating the basic reproduction number (e.g.,
\cite{hooker2010parameterizing, lin2016seasonality, pons2018serotype}). 
When we assume that the mean infectious period is known
and try to estimate the remaining 4 parameters of the SIR model, fitting the
wrong model results in clearer bias and a much lower coverage (Figure 2B); however, 
bias from fitting the wrong model changes direction (e.g., fitting the infection
model to recovery model overestimates the reproduction number instead).

@@

Mathematical modeling of infectious disease outbreaks helps us understand 
how disease spreads in a population. Disease modelers often put a lot of
effort into capturing the transmission mechanism but make naive assumptions
about how cases are reported. Here, we used a simple deterministic SIR model to show that 
making wrong assumptions about observational processes in epidemic models
can give biased estimates of the basic reproduction number and narrow
confidence intervals.

@@

While we have only considered observation processes associated with incidence 
reporting in this study, it is equally as important to consider observation
processes in mortality reporting. In particular, many disease modelers have
tried to infer underlying transmission mechanisms from historical mortality 
data but assumed that deaths are recorded as soon as individuals die 
\citep{he2013inferring, didelot2017model, dean2018human}. 
This assumption does not account for the possibility 
that a delay in reporting time of deaths can
change the shape of an epidemic curve: longer delays are expected to decrease
the size of an epidemic peak and delay the timing of the peak (Figure 1). 
These assumptions may affect their conclusions about the underlying 
transmission mechanisms.

>>>>

Epidemic models provide valuable tools for understanding how a disease
spreads in a population. Many epidemic models focus on characterizing
the underlying transmission mechanism but make naive assumptions about
how infections are reported. In this note, we use a simple deterministic
Susceptible-Infected-Removed (SIR) model to compare two popular assumptions 
about disease incidence reports: individuals can report their infection as soon as
they become infected or as soon as they recover. We show that making
wrong assumptions about the underlying observation processes can result
in biased estimates of the basic reproduction number and narrow confidence
intervals.

@@

Mechanistic analyses of epidemic time series allow us to infer the underlying 
transmission mechanism, estimate biologically relevant parameters, and 
predict the course of an outbreak \citep{breto2009time}. In order to make a 
precise and accurate inference, disease modelers try to build
more realistic process models. For example, the same London measles time 
series from the prevaccination era has been analyzed using multiple models -- 
these models account for time-varying transmission rates \citep{fine1982measles}, 
realistic age structure \citep{schenzle1984age},
metapopulation structure \citep{xia2004measles}, continuous-time infection process 
\citep{cauchemez2008likelihood}, and extra-demographic variability \citep{he2009plug}.

@@

The Susceptible-Infected-Removed (SIR) model describes how a disease spreads in a
homogeneous population:
\begin{equation}
\begin{aligned}
\frac{dS}{dt} &= - \beta S \frac{I}{N}\\
\frac{dI}{dt} &= \beta S \frac{I}{N} - \gamma I\\
\frac{dR}{dt} &= \gamma I,
\end{aligned}
\end{equation}
where $\beta$ is the contact rate per unit time, $\gamma$ is the recovery rate per unit time, 
and $N = S + I + R$ is the total population size. 
We define incidence at time $t$ as the number of newly infected
individuals that are infected between time $t- \Delta t$ and time $t$, where $\Delta t$ is
the reporting time step. We expect infected individuals to report their infection some time
after their infection; the number of reported cases defines the \emph{observed} incidence.

For brevity, we consider two extreme cases: individuals instantaneously report
their infection when they become infected or when they recover. The observed incidence 
measured upon infection, $i_1(t)$, can be defined by the integral:
\begin{equation}
i_1(t) = \int_{t - \Delta t}^{t} \beta S \frac{I}{N} dt.
\end{equation}
Alternatively, we can keep track of cumulative incidence, $C$, by adding a 
new state variable described by $dC/dt = \beta S I/N$ and taking the difference between 
the two consecutive reporting periods: $i_1(t) = C(t) - C(t-\Delta t)$. Likewise, 
incidence measured upon recovery, $i_2(t)$, can be defined by the integral:
\begin{equation}
i_2(t) = \int_{t-\Delta t}^{t} \gamma I dt,
\end{equation}
or by the consecutive difference in the cumulative number of recovered cases:
$i_2(t) = R(t) - R(t - \Delta t)$.
Finally, we model under-reporting using a negative binomial distribution with a
mean either $\rho i_1(t)$ or $\rho i_2(t)$, where $\rho$ is the reporting rate, and
an over-dispersion parameter $\theta$. For convenience, we will refer to these two
negative binomial models as infection model and recovery model hereafter; 
likewise, we will refer to epidemic time series generated from these
negative binomial models 
as infection time series and recovery time series.

@@

First, we compare the dynamics of incidence curves, $i_1(t)$ and $i_2(t)$, 
(Figure 1). A lag in reporting time delays
the observed epidemic peak timing and reduce the size of the peak. However, 
these differences in the reporting time have small effect to the overall shape 
of the epidemic curve. In the presence of observation and process error, we 
would not expect to be able to distinguish between the two reporting processes.

Note that incidence is different from prevalence,
$I(t)$, which is defined as the number of \emph{currently} infected individuals;
the dynamics of observed incidence depend on the reporting time step (because the sum of 
true incidence is equal to the final size of an epidemic), whereas those of
prevalence do not. We expect the dynamics of incidence and prevalence to be similar only when
the reporting time step is equal to the disease generation time. While it is
uncommon, some models do not make a clear distinction between the two (e.g., 
\cite{hooker2010parameterizing}).

@@

Disease modelers often assume that the mean infectious period of a disease
is known and focus on estimating the basic reproduction number (e.g.,
\cite{hooker2010parameterizing, lin2016seasonality, pons2018serotype}). 
When we assume that the mean infectious period is known
and try to estimate the remaining 4 parameters of the SIR model, fitting the
wrong model results in a clearer bias and a much lower coverage (Figure 2B); however, 
bias from fitting the wrong model changes direction (e.g., fitting the infection
model to recovery model overestimates the reproduction number instead).

@@

Mathematical modeling of infectious disease outbreaks helps us understand 
how disease spreads in a population; however, many epidemic models make 
naive assumptions about how cases are reported. 
Here, we used a deterministic SIR model to show that 
making wrong assumptions about observational processes in epidemic models
can give biased estimates of the basic reproduction number and narrow
confidence intervals.

@@

While we have only considered observation processes associated with incidence 
reporting in this study, it is equally as important to consider observation
processes in mortality reporting. In particular, many disease modelers have
tried to infer underlying transmission mechanisms from historical mortality 
data but assumed that individual deaths are recorded as soon as individuals die 
\citep{he2013inferring, didelot2017model, dean2018human}. 
This assumption does not account for the possibility 
that a delay in reporting of deaths can
change the shape of an epidemic curve: longer delays are expected to decrease
the size of an epidemic peak and delay the timing of the peak (Figure 1). 
Ignoring potential delays in reporting may affect their conclusions about the underlying 
transmission mechanisms.
